---
title: "Lab 1 Assigment"
author: "Victor S Phiri"
date: "12/8/2021"
output:
  word_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

# Load packages

```{r}
library(tidyverse)
library(psych)
library(gridExtra)
library(readr)
library(pscl) # for pR2
library(lmtest) # for lrtest
library(Amelia)
library(dominanceanalysis) # for dominanceAnalysis()
library(broom)
library(knitr)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
```

## Load the Titanic Dataset

```{r}
getwd()

Titanic_tadabase=read_csv("https://raw.githubusercontent.com/kekecsz/SIMM61-Course-materials/main/Home_assignment/Titanic%20-%20training%20set.csv")

```

## View data

```{r pressure, echo=FALSE}
str(Titanic_tadabase)
```

Record variable

```{r}
Clean_Titanic<- Titanic_tadabase %>%
  mutate(Pclass = factor(Pclass, levels = c(1, 2, 3), labels = c('First Class', 'Second Class', 'Third Class')),
	Survived = factor(Survived, levels = c(0, 1), labels = c('No', 'Yes')))
  
```

Select variables

```{r}

 Titanic_Selected <- Clean_Titanic %>% 
  select(Age,
         Sex,
         Pclass,
         Parch,
         Survived,
         SibSp)

```

# Checking the data

The data has 891 observations the Age column has missing values
representing 3% of the database

```{r}

 str(Titanic_Selected) # it will help us to understand our variables and also identify NAs 
   
#install.packages("Amelia")
 
missmap(Titanic_Selected, main = "missing values vs oberved") # assess the presence of  missing values in the database  

```

Remove NAs from the dataset We drop missing values, hence, remain with
714 observations. we use missmap to visualize our data again

```{r}
# create a new dataset without NAs called Final titanic data

Final_titanic_data<- Titanic_Selected %>% drop_na()

missmap(Final_titanic_data, main = "missing values vs oberved") # check if the all the missing values have been removed 

str(Final_titanic_data)
 summary(Final_titanic_data)
 
```

Getting details of the data

59% not survived and 41% survived

```{r}
# survival rates 

prop.table(table(Final_titanic_data$Survived))


```

Data Visualization Visualize survival rates

```{r}
 Final_titanic_data %>%
ggplot() +
aes(x = Survived) +
  geom_bar(aes(y = (..count..)/sum(..count..), fill=factor(Survived)), stat= "count")+
  geom_text(aes(label = scales::percent(round((..count..)/sum(..count..),2)),
                y= ((..count..)/sum(..count..))), stat="count",
            vjust = -.25) +
  ggtitle("Survived") + labs(y = "percent")



```

Survival Rates by Sex

The graph show high death rate among men than women

```{r}
 
Final_titanic_data %>%
ggplot() +
aes(x = Sex, fill = Survived) +
geom_bar()+
  theme_get()
labs(y="passanger count",
       title = "Titanic Survival Rates by Sex")

```

#### Titanic Survival Rates by Age by Passenger class 

Death toll was high among men in class three and also the rate of women
who died was higher than other classes

```{r}
 
 Final_titanic_data %>%
ggplot() +
aes(x = Age, fill = Survived) +
theme_dark()+
facet_wrap(Sex~Pclass)+
geom_histogram()+
labs(y="passanger count",
       tittle="Titanic Survival Rates by Age by Passanger class")
 
```

#### Class by Passenger class

the data shows that the survival rates were very low in class three as
compared to passenger classes(one and two)

```{r}

Final_titanic_data %>%
ggplot() +
aes(x = Pclass, fill = Survived) +
geom_bar()+
labs(y="passanger count",
       tittle="Titanic Survival Rates by Passanger class")
 
 
```

#### Examine survival rate by gender by passenger class 

\# data show that there were more males that perished in the third class
than other classes, \# the number of females were relatively higher in
class the upper class when compared to other classes

```{r}
Final_titanic_data %>%
ggplot() +
aes(x = Sex, fill = Survived) +
theme_bw()+
facet_wrap(~Pclass)+
geom_bar()+
labs(y="passanger count",
       tittle="Titanic Survival Rates by Gender by Passanger class") 
 
```

#### Survival rate with sibling and or spouse by class

```{r}
 
Final_titanic_data %>%
ggplot() +
aes(x = SibSp, fill = Survived) +
theme_dark()+
facet_wrap(~Pclass)+
geom_bar()+
labs(y="passanger count",
       tittle="Titanic Survival Rates by Gender by Passanger class") 
 
 
```

#### Titanic Survival chat

\
Passenger class by Parch There was a lower chance of survival for those
who were with family in the third class

```{r}
Final_titanic_data%>%
ggplot() +
aes(x = Parch, fill = Survived) +
theme_classic()+
facet_wrap(~Pclass)+
geom_bar()+
labs(y="passanger count",
       tittle="Titanic Survival Rates by Age by Passanger class by Parch")
 
```

# Design the regression model: 

The previous steps has enabled us assess our data and come up with the
following model. Outcome valuable: Survived\
Predictors: Sex, Passenger Class(Pclass), Age, Siblings or Spouse(Sibsp)
and Parent/child(Parch)

```{r}
 mod1 = glm(Survived~ Sex+Pclass+Age+SibSp+Parch,
family = binomial(), data = Final_titanic_data)

summary(mod1)

 

```

We created a table to present the coefficients clearly 

```{r}

xt<-xtable(summary(mod1))
print(xt,type = "html")%>%
  kable()

```

Calculate probabilities for Kate and Sue survival Kate Survival without
Leonard

```{r}
 # Regression equation 

Kate_Survival_Odds=4.36+(-2.64)*0+(-1.42)*0+(-2.65)*1+(-0.04)*20+(-0.37)*0+(-0.04)*1


Kate_probality= exp(Kate_Survival_Odds)/(1+ exp(Kate_Survival_Odds))*100 # probability 
 

print(Kate_Survival_Odds)
print(Kate_probality)

 
```

Kate Survival with Leonard

```{r}

Kate_Leo=4.36+(-2.64)*0+(-1.42)*0+(-2.65)*1+(-0.04)*20+(-0.37)*1+(-0.04)*1

# Odds 
Odds_Kate_Leo= exp(Kate_Leo) # survival Odds for Kate 

Kate_Leo_probality= Odds_Kate_Leo/(1+ Odds_Kate_Leo)*100  # probability for Kate with Leonard

print(Odds_Kate_Leo)
print(Kate_Leo_probality)

```

Sue survival without Leonard

```{r}
#
#Odds without her father 
Sue_SurvivalOdds = 4.36+(-2.64)*0+(-1.42)*0+(-2.65)*1+(-0.04)*4+(-0.37)*0+(-0.04)*1
 
Sue_Probality=exp(Sue_SurvivalOdds)/(1+exp(Sue_SurvivalOdds))*100
print(Sue_Probality)
 
```

Sue survival probability with Leonard on board

```{r}
Sue_Leonard = 4.36+(-2.64)*0+(-1.42)*0+(-2.65)*1+(-0.04)*4+(-0.37)*0+(-0.04)*2
 


Probalility_Sue_Leo=exp(Sue_Leonard)/(1+ exp(Sue_Leonard))*100 

print(Sue_Leonard)
print(Probalility_Sue_Leo)
```

# Model Performance

We use the R2 function to asses the performance of our model and the
null model. "pscl package".

The model has(-318.22 llh) which is bigger than the Null model(-482.26
llhNull). This implies that our model is better than the null model.

```{r}

  pR2(mod1)
head(predict(mod1))
 
```

#### Deviance

It can be seen from our data below that the our model has less
error(636.46) as compared to the Null model which has higher
deviance(964.52 llh)

```{r}
   # 2ll deviance 
pR2(mod1)["llh"] * -2 # alternative model 
 
```

#### Predicting accuracy for categorization

```{r}

Final_titanic_data = Final_titanic_data %>%
mutate(pred_mod1 = predict(mod1)) %>%
mutate(pred_mod1 = case_when(pred_mod1 <= 0.5 ~"No",
pred_mod1 > 0.5 ~ "Yes"))
 
```

#### Cording correct cases

A new variable will be added to the our final titanic dataset as shown
below;

we shall use mutate and case_when function

```{r}

 Final_titanic_data = Final_titanic_data %>%
  mutate(correct_prediction=case_when(pred_mod1 == Survived  ~ "correct",
                                      pred_mod1 != Survived ~ "incorrect"))

```

#### Check the model prediction performance

The model has 81% prediction rate(579 out of 714 cases) and 19%
incorrect predictions

```{r}
 
Final_titanic_data %>%
group_by(correct_prediction) %>%
summarise(count = n()) %>%
mutate(freq = count / sum(count))


```

# compute the null model; 

this will be used to compare the data with the data predicted by our
model.

  

```{r}
mod_null = glm(Survived ~ 1, family = binomial(), data = Final_titanic_data)

summary(mod_null)

```


The null model

```{r}

Final_titanic_data %>%
group_by(Survived) %>%
summarise(count = n()) %>%
mutate(freq = count / sum(count))
 
```

compare the predictions of the null model with the alternative model

```{r}

Final_titanic_data%>%
group_by(Survived, pred_mod1) %>%
summarize(n = n()) %>%
spread(Survived, n)


```

### check for the prediction of Survival rate by response category(Survived= Yes/No)

63% of survivors were correctly predicted. However, 36% were wrongly
predicted. Further this prediction does not meet our required target of
72%.

```{r}

Final_titanic_data %>%
filter(Survived == "Yes") %>%
group_by(correct_prediction) %>%
summarise(count = n()) %>%
mutate(freq = count / sum(count))

```

#### Predictions for those who died

our model had 92% correct prediction and 7% incorrect predictions

```{r}


Final_titanic_data %>%
filter(Survived == "No") %>%
group_by(correct_prediction) %>%
summarise(count = n()) %>%
mutate(freq = count / sum(count))

```

# Adjusting our model
Since our model did not give meet the required minimum prediction of 72% for for both cases, 

We adjusted our model by changing the cut off point from 0.5 to 0

```{r}
  
Final_titanic_data = Final_titanic_data %>%
mutate(pred_mod1_ajusted = predict(mod1)) %>%
mutate(pred_mod1_ajusted = case_when(pred_mod1_ajusted <= 0 ~"No",
pred_mod1_ajusted > 0 ~ "Yes"))
 
```

# coding correct guesses

```{r}
 
Final_titanic_data = Final_titanic_data %>%
  mutate(correct_prediction_Ajusted=case_when(pred_mod1_ajusted == Survived  ~ "correct",
                                      pred_mod1_ajusted != Survived ~ "incorrect"))
 
```

# correct categorization rate overall

 The adjusted model has 81% overall observations 

```{r message=FALSE, results='asis'}
  
Final_titanic_data %>%
group_by(correct_prediction_Ajusted) %>%
summarise(count = n()) %>%
mutate(freq = count / sum(count))
 
```

# correctly categorized as Survived(Yes)

73% of the cases were correctly predicted 

```{r}

Final_titanic_data %>%
filter(Survived == "Yes") %>%
group_by(correct_prediction_Ajusted) %>%
summarise(count = n()) %>%
mutate(freq = count / sum(count))

```

# correctly categorized as Not Survived(No)

86% were correctly predicted 

```{r}

 Final_titanic_data %>%
filter(Survived == "No") %>%
group_by(correct_prediction_Ajusted) %>%
summarise(count = n()) %>%
mutate(freq = count / sum(count))
 
```

#Comparing our model with the null model 

The data shows that our model is significantly better than the null model. 
with log of likelihood -318.23 which is better than -482.26(Null model). Further our model has a significant Pvalue as shown below 

```{r}
 lrtest(mod_null, mod1)
 
```
#model fit 

The AIC for the model is 650.46 which is lower than 966.52 for the null model as shown in the output below 

```{r}

 AIC(mod_null, mod1)
 
```

```{r}
exp(coef(mod1))

```

#Dominance Analysis 

```{r message=FALSE, results='asis'}

Dominance_Mod1<-dominanceAnalysis(mod1)

contributionByLevel(Dominance_Mod1, fit.functions="r2.m")

```
# Average contribution of predictors 

We conclude from our findings that Sex(0.21) is the main predictor of Survival outcome, followed by Passenger class(0.11), Age(0.02), Sibling or spouse(0.01), and parent or child(0.00)

note(above figures have been round to two decimal places)

```{r}

averageContribution(Dominance_Mod1,fit.functions = "r2.m")
   

```

The 


```{r}

plot(Dominance_Mod1, which.graph ="conditional",fit.function = "r2.m")



```



```{r}
plot(Dominance_Mod1, which.graph ="general",fit.function = "r2.m")


```
Presentation of summary table 

```{r}
 library(stargazer)


tab_model(mod1,show.se =  FALSE, show.std = TRUE, show.stat = TRUE,show.est = TRUE, 
           show.aic = TRUE,show.dev = TRUE, 
           show.loglik=TRUE)


```


```{r}


```

