---
title: "Assigment two, mixed models"
author: "Victor S Phiri"
date: "1/4/2022"
output: word_document
output:output:
  pdf_document: null
  html_document: "word"
---

# Lord packages {#lord-packages}

```{r message=FALSE, paged.print=TRUE}
 
library(psych) # for describe	
library(tidyverse) # for tidy code and ggplot		
library(lme4) # for lmer() mixed models	
library(lmerTest)  # for significance test on lmer() mixed models	
library(cAIC4) # for cAIC	
library(r2glmm) # for r2beta	
library(MuMIn) # for r.squaredGLMM
library(kableExtra) #for tables 
library(sjPlot)
library("Amelia") # checking NAs
library(gridExtra)
```

## \# for extraction of standardised beta coefficients

```{r cars}
  # for extration of standarzed beta coefficients 
 stdCoef.merMod <- function(object) {	
  sdy <- sd(getME(object,"y"))	
  sdx <- apply(getME(object,"X"), 2, sd)	
  sc <- fixef(object)*sdx/sdy	
  se.fixef <- coef(summary(object))[,"Std. Error"]	
  se <- se.fixef*sdx/sdy	
  return(data.frame(stdcoef=sc, stdse=se))	
}	


```

## Load datasets

```{r}
library(readr)
surgery_data_A <- read_csv("C:/Users/Saidi/OneDrive/Desktop/Logistic Regression/surgery_data_A.csv")

view(surgery_data_A)#view data
str(surgery_data_A)#data structure 
 
surgery_data_B <- read_csv("C:/Users/Saidi/OneDrive/Desktop/Logistic Regression/surgery_data_B.csv")
view(surgery_data_B)
str(surgery_data_B)

```

\#Check for missing values no missing values in our databases

```{r}
missmap(surgery_data_A, main = "missing values vs oberved")
missmap(surgery_data_B, main = "missing values vs oberved")
```

# \#Record variables

Convert character variables into factors hospital, Sex and ID which are character but variables

```{r}
# record variables factors 

surgery_data_A=surgery_data_A%>%
  mutate(ID=as_factor(ID),
         sex=as_factor(sex),
         hospital=as_factor(hospital))

 str(surgery_data_A)

surgery_data_B=surgery_data_B%>%
  mutate(ID=as_factor(ID),
         sex=as_factor(sex),
         hospital=as_factor(hospital))



```

## \#Recod2 variables

Sex in database A has an error- 3 levels male,female,woman; we recode it into two levels,male and female

```{r}
surgery_data_A <- surgery_data_A %>% 
        mutate(sex = recode(sex, 
                            male=c("male"),female=c("female"),woman=c("female")))
                            
                                 
                                       
str(surgery_data_A$sex)




```

# View Descriptive statistics of data A

```{r echo=FALSE}
describe(surgery_data_A)
 

```

## **visualize the data**

age has a negative correlation with pain while cortisol serum has a strong positive relation with pain.

```{r}
# pain- age relationship
figure1=surgery_data_A %>% 		
  ggplot() +		
  aes(y = pain, x = age) +		
  geom_point(aes(color = sex), size = 3) +
  
  geom_smooth(method = "lm", se = F)

# pain - cortisol_serum relationship
figure2=surgery_data_A %>% 		
  ggplot() +		
  aes(y = pain, x = cortisol_serum) +		
  geom_point(aes(color = sex), size = 3) +
  
  geom_smooth(method = "lm", se = F)


grid.arrange(figure1,figure2,nrow=2)


```

**\# check the relationship of pain with cortisol_serum in each hospital**

```{r echo=TRUE}
# check the relationship of pain with cortisol_serum in each hospital 

surgery_data_A%>%
ggplot() +
 aes(y = pain, x = cortisol_serum) +		
  geom_point(aes(color = sex), size = 5) +
theme_bw()+
facet_wrap(~hospital)+
  geom_smooth(method = "lm", se = F)
 
```

### Create Random Intercept model

we shall call our model, mod_RI

```{r echo=TRUE}

mod_RI<- lmer(pain~sex+age+cortisol_serum+ STAI_trait+pain_cat+mindfulness+(1|hospital), data = surgery_data_A)


summary(mod_RI)
```

## Confidence intervals

```{r echo=FALSE}
confint(mod_RI)
```

#### Model Fit

**Marginal R2**

To check for R2 for our Fixed effect predictors, we used the`R2beta`. our output shows that cortisol_serum and pain_cat has higher influence at 95% confidence interval than than other variables. Standardised beta for each predictor was also noted but the conclusion was as above.

```{r echo=TRUE}
r2beta(mod_RI,method ="nsj",data =surgery_data_A)

 
```

using this method, our fixed effect predictors explained 39% of the variance while the while the model explained 46% of the variance.

```{r echo=TRUE}

r.squaredGLMM(mod_RI)


```

## conditional R2

```{r echo=TRUE}
 
cAIC(mod_RI)$caic

```

Model variance

Null model

```{r echo=TRUE}
mod_mean<- lm(pain~1,surgery_data_A)
mod_mean
```

\#Predict database B

we create a new object for predicted values

```{r}
pred_DataB<-predict(mod_RI,newdata=surgery_data_B,allow.new.levels=TRUE)
```

Compute TSS

```{r}
TSS=sum((surgery_data_B$pain-predict(mod_mean))^2)

print(TSS) 

```

Compute RSS

```{r}
RSS=sum((surgery_data_B$pain -pred_DataB)^2)

print(RSS) 
```

```{r}

R2=1-(RSS/TSS)

R2

```

\#New Model

Random slop model cortisol_serum was noted in our previous model as the most influential predictor

Random Slop model

```{r}
new.Rand_Int<-lmer(pain ~ cortisol_serum + (1|hospital),data = surgery_data_A)
new.Rand_Int

Mod2_RS=lmer(pain ~ cortisol_serum  + ( cortisol_serum|hospital),data = surgery_data_A)
Mod2_RS



```

#check model fit 
we compare the model fit of our two models 

```{r}
sum(residuals(new.Rand_Int)^2)
sum(residuals(Mod2_RS)^2)
```

```{r echo=TRUE}

cAIC(new.Rand_Int)$caic

cAIC(Mod2_RS)$caic

anova(Mod2_RS,new.Rand_Int)

```



Add predicted values to b

```{r}
 
surgery_data_A=surgery_data_A%>%
mutate(pred_int =predict(mod_RI),pred_slope =predict(Mod2_RS),pred_new_rand_int=predict(new.Rand_Int))


```

Random Slop random slop has a better fit than the random intercept as seen below.

```{r echo=TRUE, fig.height=6, fig.width=10}

# random slope

graph1_RS<-ggplot(surgery_data_A,aes(y = pain,x =cortisol_serum ,
group =hospital))+
  geom_point(aes(colour=hospital,size =4))+geom_line(color ="red",size=1,
aes(y =pred_slope,x =cortisol_serum ))+
  facet_wrap(~hospital,ncol =5)+
  theme_gray()
  
graph1_rand_int<- ggplot(surgery_data_A,aes(y = pain,x =cortisol_serum ,
group =hospital))+
  geom_point(aes(colour=hospital,size =4))+
  geom_line(color ="red",size=1,
aes(y =pred_new_rand_int,x =cortisol_serum ))+
  facet_wrap(~hospital,ncol =5)+
  theme_gray()


grid.arrange(graph1_rand_int,graph1_RS,nrow=2)

```



Generate table for the report

```{r eval=FALSE, message=FALSE, include=FALSE, results='asis'}

tab_model(mod_RI,CSS = list(css.depvarhead = '+color: red;',css.firsttablecol = 'font-weight: bold;',css.summary = 'color: blue;'))
```
